---
title: 'Gerrymandering Analysis: 1790-1860'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(plyr)
library(dplyr)
library(ggmap)
library(stringi)
library(nlstools)
library(minpack.lm)
library(drc)
require(forcats)
data <- read.csv("output_with_incumbency.csv")
```

## Introduction

Gerrymandering is a hot topic in statehouses, academia and the Supreme Court. It was named for Elbridge Gerry, Massachusetts Governor from 1810-1812 and alleged gerrymandering mastermind, but as far as I know not much study has been done on gerrymandering in the pre-Civil War era. This analysis aims to scratch the surface and provide a glimpse of the prevalence of gerrymandering in that early stage in our history.


## Summaries
Parties Represented:
```{r}
data$party <- as.character(data$party)
data[,"party"][data[,"party"] == "D-R"] <- "DR"
data <- separate(data, party, into=c("party", "subparty"), sep="-", fill="right")
data[,"party"][data[,"party"] == "Ad"] <- "NR"
data[,"party"][data[,"party"] == "f"] <- "F"
data[,"party"][data[,"party"] == "a-f"] <- "AF"
data[,"party"][data[,"party"] == "J"] <- "D"
table(data$party)
```

```{r}
ggplot(data, aes(x=percentage)) + 
  geom_histogram() + 
  ggtitle("Distribution of vote percentages won by candidates")

ggplot(data, aes(x=num_elected)) + 
  geom_histogram() + 
  ggtitle("Distribution of number of representatives elected per district")

party_summary <- data %>%
  group_by(party) %>%
  summarise(won=sum(result=="won", na.rm=TRUE)) %>%
  filter(won > sum(data$result == "won", na.rm=TRUE)*.1/8)

ggplot(party_summary, aes(x=reorder(party, -won), y = won)) + 
  geom_bar(stat="identity") + 
  ggtitle("Major Parties Represented") + 
  xlab("Party") + 
  ylab("Number of Wins")
```
Party Key:

D: Democrat

DR: Democratic-Republican

W: Whig

F: Federalist

R: Republican

(Blank): No party

NR: National Republican (also called Anti-Jacksonian)

U: Union

A: American (also known as Know-Nothing)


```{r}
data <- transform(data,
                  election_id = as.numeric(interaction(congress, trial,
                                                       year, state,
                                                       district, drop=TRUE)))

data$congress <- as.factor(data$congress)
summary <- data %>% 
  group_by(election_id) %>% 
  summarise(win_percentage=max(percentage))

ggplot(summary, aes(x=win_percentage)) + 
  geom_histogram() + 
  ggtitle("Win Percentage Distribution")

data <- data %>% 
  group_by(election_id) %>% 
  mutate(inferred_percentage=ifelse(!is.na(percentage) | num_elected > 1,
                                    percentage, 
                                    votes/sum(votes)*100*num_elected))

```
```{r}
data <- transform(data, state_id = as.numeric(interaction(congress, state, drop=TRUE)))

to_join <- data %>% 
  group_by(state_id) %>%
  summarise(num_elected_state=sum(result=="won"))

data_joined <- inner_join(data, to_join, by="state_id")

grouped_states <- data_joined %>% group_by(state, congress) %>% 
  mutate(sum_votes = sum(votes * (type=="StandardElections"),
                         na.rm=TRUE)) %>% 
  group_by(state, congress) %>% 
  mutate(num_districts = length(unique(district))) %>%
  group_by(state, congress, party) %>%
  summarise(theoretical = sum(votes * (type=="StandardElections"), 
                              na.rm=TRUE)/mean(sum_votes),
            actual=sum(result=="won")/max(num_elected_state),
            num_elected_state=max(num_elected_state), 
            num_districts = max(num_districts)) %>%
  mutate(x= theoretical, theoretical = (theoretical-.5)*2+.5) %>%
  mutate(theoretical = ifelse(theoretical > 1, 1, 
                              ifelse(theoretical < 0, 0, theoretical)),
         mean_elected_district=num_elected_state/num_districts)
```
## Calculating the Efficiency Gap
First, I use the efficiency gap calculation proposed by Nicholas Stephanopoulos and Eric McGhee, specifically the simplified version described [here](https://www.brennancenter.org/sites/default/files/legal-work/How_the_Efficiency_Gap_Standard_Works.pdf). I calculate the efficiency gap for a given state, then floor it so that it represents a whole number of seats. For example, if the efficiency gap is 0.27 for a state with 4 seats, that gap represents 1.08 seats. This number is floored to one seat, to make a floored gap of 0.25. 

```{r}
# party %in% c("D-R", "F", "J", "NR", "Ad")
filtered <- subset(grouped_states, party != "" & 
                     !is.na(theoretical) & 
                     !is.na(actual) & 
                     num_districts>0)

filtered$sign <- ifelse(filtered$actual-filtered$theoretical > 0, 1, -1)
filtered$simple_efficiency_gap <- filtered$sign * 
  floor(abs(filtered$actual-filtered$theoretical)*filtered$num_elected_state)/
  filtered$num_elected_state
ggplot(filtered, aes(x=simple_efficiency_gap)) + 
  geom_histogram() + 
  ggtitle("Distribution of Simple Adjusted Efficiency Gaps")
```

Below is a plot of comparing the percentages of seats won to the percentage of votes won, along with the prediction from the basic model.

```{r}
ggplot(filtered, aes(x=x, y=actual)) + 
  geom_point(aes(alpha=0.001)) + 
  xlab("Percentage of Votes Won") + 
  ylab("Percentage of Seats Won") + 
  guides(alpha=FALSE) + 
  geom_line(aes(x=x, y=theoretical)) + 
  ggtitle("Seats Vs. Votes in Theory (line) and in Actuality (points)")
```

Next, I filter the data to include only frequent parties, and collect the data by party, state, and census.

```{r}
filtered <- filtered %>% 
  filter(party %in% c("D", "DR", "W", "F", "R", "NR","U", "A"))
data$census <- floor((as.integer(data$congress)+2)/5)
filtered$census <- floor((as.integer(filtered$congress)+2)/5)
filtered$seat_advantage <- filtered$simple_efficiency_gap*filtered$num_elected_state
```
In my observation, it appears that states with fewer districts typically have seat percentage outcomes that are closer to zero or one. I build a new model predicting seat percentage, taking this into account, and calculate efficiency gaps (difference between predicted and actual seat share) for both the traditional and modified predictions.
```{r}
model <- lm(abs(filtered$actual-.5) ~ abs(filtered$x-.5) + log(filtered$num_districts))
summary(model)
new_predictions <- .5+ifelse(filtered$x < .5, -1, 1)*predict(model)
new_predictions <- ifelse(new_predictions < 0, 0, ifelse(new_predictions > 1, 1, new_predictions))
filtered$theoretical_2 <- new_predictions
model <- lm(actual ~ 0 + theoretical_2, filtered)
summary(model)
summary(lm(actual ~ 0 + theoretical, filtered))
filtered$sign <- ifelse(filtered$actual-filtered$theoretical_2 > 0, 1, -1)
filtered$modified_efficiency_gap <- filtered$sign * 
  floor(abs(filtered$actual-filtered$theoretical_2)*filtered$num_elected_state)/
  filtered$num_elected_state

state_profile <- filtered %>% 
  group_by(party, census, state) %>% 
  summarise(mean_simple_gap=mean(simple_efficiency_gap, na.rm=TRUE),
            mean_modified_gap=mean(modified_efficiency_gap, na.rm=T),
            theoretical=mean(theoretical, na.rm=TRUE),
            actual=mean(actual, na.rm=TRUE),
            num_districts=mean(num_districts),
            seat_advantage=mean(seat_advantage))

write.csv(data_joined, "data_joined.csv")
```

```{r}
states <- map_data("state")
state_profile$region <- sapply(state_profile$state, tolower)
write.csv(filtered, "filtered.csv")
```

Below are maps representing the efficiency gaps from each party present at each census:

```{r}
for (this_census in 1:7) {
  for (this_party in unique(state_profile$party)) {
    states_with_data <- left_join(states,
                                  subset(state_profile, census==this_census & party==this_party),
                                  by = "region")
    if (sum(!is.na(states_with_data$mean_modified_gap)) > 0) {
      plot.new()
      print(
        ggplot(data = states_with_data) + 
          geom_polygon(aes(x = long, y = lat,
                           group = group, fill=mean_simple_gap),
                       color = "white") + 
          coord_fixed(1.3) +
          scale_fill_gradient2(high="#5ab4ac", low="#d8b365") + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.line=element_blank(),
                axis.text.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks=element_blank()) +
          xlab("") +
          ylab("") + 
          labs(fill= this_party %s+% " Advantage") + 
          ggtitle("Census #" %s+% this_census %s+% " " %s+% this_party %s+% ": simple")
        )
      print(
        ggplot(data = states_with_data) + 
          geom_polygon(aes(x = long, y = lat,
                           group = group, fill=mean_modified_gap),
                       color = "white") + 
          coord_fixed(1.3) +
          scale_fill_gradient2(high="#5ab4ac", low="#d8b365") + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.line=element_blank(),
                axis.text.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks=element_blank()) +
          xlab("") +
          ylab("") + 
          labs(fill= this_party %s+% " Advantage") + 
          ggtitle("Census #" %s+% this_census %s+% " " %s+% this_party  %s+% ": modified")
        )
    }
  }
}
```

Below are similar maps, but representing the mean number of extra seats the party won as a result of the gap:

```{r}
for (this_census in 1:7) {
  for (this_party in unique(state_profile$party)) {
    states_with_data <- left_join(states,
                                  subset(state_profile, census==this_census & party==this_party),
                                  by = "region")
    if (sum(!is.na(states_with_data$seat_advantage)) > 0) {
      plot.new()
      print(
        ggplot(data = states_with_data) + 
          geom_polygon(aes(x = long, y = lat, 
                           group = group, fill=seat_advantage),
                       color = "white") + 
          coord_fixed(1.3) +
          scale_fill_gradient2(high="#5ab4ac", low="#d8b365") + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.line=element_blank(),
                axis.text.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks=element_blank()) +
          xlab("") + 
          ylab("") +
          labs(fill= this_party %s+% " Mean Seat Advantage") + 
          ggtitle("Census #" %s+% this_census %s+% " " %s+% this_party)
        )
    }
  }
}
```